---
title: Getting and Cleaning the *Human Activity Recognition Using Smartphones Database*
  - Course Project
author: "Angie Marchany-Rivera"
date: "7/16/2020"
output:
  html_document: default
  pdf_document: default
---
Click this link to access the HTML version of the README.rmd file: https://angierosy02.github.io/GettingAndCleaningDataProject/README.html

## Description
The purpose of this project is to create a tidy data set based on the Human Activity Recognition Using Smartphones Database available at the [UCI Machine Learning Reporsitory](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones) that shows the average for each of the features and subjects. Access the \"codebook\_tidyData.rmd\" file to read detailed descriptions for each variable used in this data set.

You can download the data for this project from the following link:
https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip   

The data set linked above represents data collected from the accelerometers from the Samsung Galaxy S smartphone. This dataset contains the following files:  

1. activity\_labels.txt \- Includes the identifier and the name of each activity performed by the subject  
2. features.txt \- Includes the variable names for all 561 features  
3. features_info.txt \- Information about how the variables were measured and analyzed  
4. README.txt \- Information about the experiment, data collection and analysis  
5. test folder \- The test data is located in the \"X\_test\.txt\", the activity identifier is located in the \"Y\_test\.txt\" and the subject identifier for each row of data is located in the \"subject\_test\.txt\". Other files in this folder were not used in this project.  
6. train folder \- The train data is located in the \"X\_train\.txt\", the activity identifier is located in the \"Y\_train\.txt\" and the subject identifier for each row of data is located in the \"subject\_train\.txt\". Other files in this folder were not used in this project.  



## Code Description
Note: The following code can be found in the \"run_analysis.R\" file.

### Installed Packages  
```{r, eval = FALSE}
install.packages("dplyr")  
install.packages("gsubfn")  
install.packages("tidyr")
```

```{r, message = FALSE, warning = FALSE}
library(dplyr)  
library(gsubfn)  
library(tidyr)
```

### Getting the Data from the database link and unzipping the folder into the local directory
```{r, message = FALSE, warning = FALSE}
setwd("C:/Users/angie/Desktop/GettingAndCleaningDataProject")

if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(fileUrl,destfile="./data/Dataset.zip",method="curl")
unzip(zipfile="./data/Dataset.zip",exdir="./data")
```

### Finding the file path and saving the file names into a list
```{r, message = FALSE, warning = FALSE}
path_rf <- file.path("./data" , "UCI HAR Dataset")
list_files<- list.files(path_rf, full.names = T, recursive = TRUE)
list_files
```

### Reading the data from the targeted files.
Review the output of ```list_files``` as reference for the indexes used below.  

```{r, message = FALSE, warning = FALSE}
activity_labels <-read.table(list_files[[1]], header = F, col.names = c("id","Activity"),stringsAsFactors = TRUE) #labels for each activity
col_names <- read.table(list_files[[2]], header = F) #List of features

# Column bind the subject and activity identifiers into the first and second column of the test data frame
test_data <- cbind(read.table(list_files[[16]], header = F, col.names = "Activity"),
                   read.table(list_files[[14]], header = F, col.names = "Subject"),
                   read.table(list_files[[15]], header = F, col.names = col_names[,2]))

# Column bind the subject and activity identifiers into the first and second column of the train data frame
train_data <- cbind(read.table(list_files[[28]], header = F,col.names = "Activity"),
                    read.table(list_files[[26]], header = F, col.names = "Subject"),
                    read.table(list_files[[27]], header = F, col.names = col_names[,2]))
```

### Merging the ```train_data``` and ```test_data``` data frames into one data frame using the row binding function ```rbind```.

```{r, message = FALSE, warning = FALSE}
allData <- rbind(test_data,train_data)

#Sample output of allData. Notice that there are no descriptive names for the activities or features (i.e, column names) in this data frame.
allData[1:2,1:20] 
```
### Using descriptive activity names to replace the activity id in the ```allData``` data frame.

```{r, message = FALSE, warning = FALSE}
#Make the labels descriptive by removing the "_" in between words.
activity_labels <- gsub("_"," ",activity_labels$Activity)

# Replace the activity id number with its corresponding label.
allData$Activity<-factor(allData$Activity, labels = activity_labels)

#Sample output of allData. Notice that the activities are now labeled.
allData[1:2,1:20]
```
### Extracting only the measurements on the mean and standard deviation for each feature in the ```allData``` data frame.

```{r, message = FALSE, warning = FALSE}
subData <- allData %>% select(Activity, Subject, contains(".mean."),contains(".std."))

#Sample output of subData.
subData[1:2,1:20]
```
### Creating a second, independent tidy data set with the average of each feature for each activity and each subject in the ```subData``` data frame.

```{r, message = FALSE, warning = FALSE}
tidyData <- subData %>% group_by(Subject, Activity) %>% 
        summarize(across(.cols = everything(),.fns = mean)) %>%
        gather(key = "Feature", value = "Average", - Activity,- Subject)

#Sample output of tidyData.
head(tidyData, n=4)
```

### Labeling the data set with descriptive variable names by separating the parts of each feature into a unique column.
```{r, message = FALSE, warning = FALSE}
 # Creates a character vector with the patterns that show up on the name of each feature.
patterns <- c("^t","^f","(Body){1}|(Body){2}","Acc","Gyro","Jerk","Gravity","Mag","\\.mean|\\.mean\\.\\.",
              "\\.std|\\.std\\.\\.","\\.X|\\.\\.\\.X","\\.Y|\\.\\.\\.Y","\\.Z|\\.\\.\\.Z")

 # Creates a character vector with the replacement for each pattern that shows up on the name of each feature.
replacements <-c("Time","Frequency","_Body","_Accelerometer","_Gyroscope",
                 " Jerk","_Gravity"," Magnitude","_Mean","_std"," in X direction"," in Y direction"," in Z direction")

# Apply the gsub function to find each pattern and replace it.
for(i in 1:length(patterns)){
        tidyData$Feature <- gsub(patterns[i],replacements[i],tidyData$Feature)}

#Separate each word on the feature's name and assign it to a new column with descriptive variable names.
tidyData <- tidyData %>%  separate(Feature, c("Domain","Motion","Sensor","Values"),sep = "_")

#Final output
tidyData
```

To print ```tidyData``` into a text file use:
```{r, eval=FALSE}
write.table(tidyData,file = "tidyData.txt", sep = "\t", row.names = FALSE)
```
## Codebook for tidyData: \"codebook_tidyData.rmd\"

The codebook for ```tidyData``` was generated using the dataMaid package:  ```library(dataMaid), makeCodebook(tidyData)```. This codebook was edited to show labels and descriptions for each variable in the ```tidyData``` data frame.

## Acknowledgment
This project was completed as part of the Getting and Cleaning Data Course from John Hopkins University Data Science Specialication in Coursera.

